A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-14 paper/paper.tex
Add text to related work

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We will use
the unscoped de Bruijn representation for the rest of the paper,
adopting notations from \citet{autosubst2}, summarized in Figure~\ref{fig:auxdef}.
Without providing
the full definition of the renaming and substitution functions, it is
impossible to tell the binding structure from the syntax
alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
with the de Bruijn depth of each term, though we note that the
syntax we work with is unscoped and the choice does matter when
we extend our logical relation to open terms in Section~\ref{sec:discuss}.

@@

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT, where the judgmental equality typically
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. Furthermore, it is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types) and we discuss in Section~\ref{sec:discuss} how we can extend
our proof to handle those rules.

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity described
in \citet{siles2012pure}. Section~\ref{sec:discuss} explains how we
can generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

Finally, since our system has an infinite universe hierarchy, we can
present the system à la Russell by using the same judgment form $[[G
|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
is therefore no need to distinguish between big types and small types
and duplicate our typing specification.

@@

The automation here not only gives us a proof that is shorter and more
resiliant to changes in definition, but also gives useful
documentation for readers who wish to understand how the underlying
proof works since the automation tactic is guaranteed to not use
lemmas or invert derivations that are not specified in the
\texttt{use} or \texttt{inv} flag.
\section{Discussion}
\label{sec:discuss}

\subsection{Use of Impredicativity}

\section{Related Work}
\label{sec:relatedwork}

>>>>

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We will use
the unscoped de Bruijn representation for the rest of the paper,
adopting notations from \citet{autosubst2}, summarized in Figure~\ref{fig:auxdef}.
Without providing
the full definition of the renaming and substitution functions, it is
impossible to tell the binding structure from the syntax
alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
with the de Bruijn depth of each term, though we note that the
syntax we work with is unscoped and the choice does matter when
we extend our logical relation to open terms in Section~\ref{sec:extension}.

@@

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT, where the judgmental equality typically
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. Furthermore, it is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types). % and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity described
in \citet{siles2012pure}. Section~\ref{sec:extension} explains how we
generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

Finally, since our system has an infinite universe hierarchy, we can
present the system à la Russell by using the same judgment form $[[G
|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
is no need to distinguish between big types and small types
and duplicate our typing specification.

@@

The automation here not only gives us a proof that is shorter and more
resiliant to changes in definition, but also gives useful
documentation for readers who wish to understand how the underlying
proof works since the automation tactic is guaranteed to not use
lemmas or invert derivations that are not specified in the
\texttt{use} or \texttt{inv} flag.

\section{Related Work}
\label{sec:relatedwork}
\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that establish metatheoretic results strong enough to derive
consistency for dependently typed systems. Some of these techniques
are still in use in more recent works. For example, the technique for
proving strong normalization by \citet{geuvers1994short} is adopted by
\citet{moon2021graded} to show the same property to a dependently
typed system extended with modalities.

More recently, there have been efforts at mechanizing consistency
or decidability of type conversion proofs using proof assistants such
as Coq and Agda.
